Convective heat transport in stratified atmospheres at low and high Mach number

### 2016-10-27 tex/nusselt_paper.tex
reworks experiment section.  Cleans it up?

\section{Experiment} 
\refstepcounter{section}
\label{sec:experiment}
In order to compare our results with previous studies and in an effort to examine a simplest case,
we study a fluid composed of monatomic ideal gas particles with an adiabatic index of $\gamma = 5/3$ and
whose equation of state is $P = R\rho T$. In constructing an initial atmosphere, we assume
that the gravitational
acceleration and conductive heat flux do not vary with depth. Both 
the thermal conductivity, $\kappa$, and the temperature gradient, $\grad T_0$, are constant,
such that $\bm{F}_{\text{cond,0}} = -\kappa \grad T_0 = \text{constant}$.
Under these assumptions, satisfying hydrostatic equilibrium produces an atmosphere defined by
\begin{equation}
\begin{split}
\rho_0(z) &= \rho_{t}(z_0 - z)^m \\
T_0(z)    &= T_{t}(z_0 - z).
\label{eqn:polytrope}
\end{split}
\end{equation}
The polytropic index is $m = m_{ad} - \epsilon$
where $m_{ad} \equiv (\gamma-1)^{-1}$ is the adiabatic polytropic index and $\epsilon$ is the
superadiabatic excess.
The subsequent entropy gradient at the top of the atmosphere is $\grad S(L_z) = -\epsilon$. 
Thermodynamic variables are nondimensionalized at the top of the atmosphere as 
$P_0(L_z) = \rho_0(L_z) = T_0(L_z) = 1$, requiring $z_0 \equiv L_z + 1$ and $R = T_{t} = \rho_{t} = 1$.
By this choice, the non-dimensional length scale is the inverse temperature gradient scale and
the timescale is the isothermal sound crossing time of this unit length.
$z$ increases upwards within $[0, L_{z}]$, where $L_{z} = e^{n_{\rho}/m} - 1$ is
determined by the number of density scale heights the atmosphere spans, $n_\rho$.
Throughout this letter, we set $n_{\rho} = 3$ such that the density varies by a factor of 20.
The characteristic timescale of convective dynamics
is related to the atmospheric buoyancy time, $t_{\text{b}} = \sqrt{L_z/g\epsilon}$, with $g = (m+1)$.
We use buoyancy time units in this letter.

Atmospheric diffusivities are specified by the Rayleigh number and the Prandtl number.  The
non-dimensional Rayleigh number is
\begin{equation}
\text{Ra} = \frac{g L_z^3 (\Delta S_0 / c_P)}{\nu\chi},
\end{equation}
where $\Delta S_0$ is the entropy difference between the top and bottom of the atmosphere, 
$c_P = R\gamma(\gamma-1)^{-1}$ is the specific heat at constant pressure,
$\nu$ is the kinematic viscosity (viscous diffusivity), and $\chi$ is the thermal diffusivity.  
The relationship between the thermal and viscous diffusivities is
set by the Prandtl number, Pr$ = \nu/\chi$.   We relate the dynamic viscosity, $\mu$, and the thermal conductivity,
$\kappa$, to their corresponding diffusivities such that 
$\nu \equiv \mu/\rho$ and $\chi \equiv \kappa/\rho$.  As a result, $\text{Ra} \propto (\nu\chi)^{-1} \propto
\rho^2$.  The atmospheres studied here with $n_{\rho} = 3$ experience an increase in the Rayleigh number 
by a factor of 400 across the domain.  This formulation leaves Pr
constant throughout the depth of the atmosphere. In this letter we impose $\text{Pr} = 1$ and
specify Ra at $z = L_z$.  We study atmospheres with an aspect ratio of 4.

@@

The Fully Compressible Navier-Stokes equations,
\begin{align}
&\begin{aligned}
&\frac{\partial \ln\rho}{\partial t} + \Div{\bm{u}} = -\bm{u}\cdot\grad\rho,
	\label{eqn:continuity_eqn}
\end{aligned}\\
&\begin{aligned}
\frac{\partial\bm{u}}{\partial t} + \grad T - &\nu\Div{\lilstressT} - \lilstressT\cdot\grad\nu = \\
&-\bm{u}\cdot\grad\bm{u} - T\grad\ln\rho + \bm{g} + \nu\lilstressT\cdot\grad\ln\rho,
\label{eqn:momentum_eqn}
\end{aligned}\\
&\begin{aligned}
\frac{\partial T}{\partial t} -\frac{1}{c_V}\left(\right.\chi&\left.\grad^2 T + \grad T\cdot\grad\chi\right) = \\
	&-\bm{u}\cdot\grad T - (\gamma-1)T\Div{\bm{u}} \\
	&+ \frac{1}{c_V}\left(\chi\grad T \cdot\grad\ln\rho +\nu\left[\lilstressT\cdot\nabla\right]\cdot\bm{u}\right), 
	\label{eqn:energy_eqn}
\end{aligned}
\end{align}
are evolved with a viscous stress tensor defined as
\begin{equation}
\sigma_{ij} \equiv \left(\frac{\partial u_i}{\partial x_j} + \frac{\partial u_j}{\partial x_i} - \frac{2}{3}\delta_{ij}\Div{\bm{u}}\right).
	\label{eqn:stress_tensor}
\end{equation}
These equations carry a total convective flux of
\begin{equation}
\bm{F}_{\text{conv}} \equiv \bm{F}_{\text{enth}} + \bm{F}_{\text{KE}} + \bm{F}_{\text{PE}} + \bm{F}_{\text{visc}},
\end{equation}
where $\bm{F}_{\text{enth}} \equiv \rho\bm{u}(c_V T + P/\rho)$ is the enthalpy flux, $\bm{F}_{\text{KE}} \equiv 
\rho|\bm{u}|^2\bm{u}/2$ is the kinetic energy flux, $\bm{F}_{\text{PE}} \equiv \rho\bm{u}\phi$ is the potential
energy flux (with $\phi \equiv -gz$), 
and $\bm{F}_{\text{visc}} \equiv -\rho\nu\bm{u}\cdot\lilstressT$ is the viscous flux. 
%Taking an inner product of
%Eq. \ref{eqn:momentum_eqn} with $\bm{u}$ and adding it to 
%Eq. \ref{eqn:energy_eqn}, the full energy equation in conservation form is retrieved,
%\begin{equation}
%\frac{\partial}{\partial t}\left(\rho\left[\frac{|\bm{u}|^2}{2} + c_V T + \phi\right]\right) +
%\Div{\bm{F}_{\text{conv}} + \bm{F}_{\text{cond}}} = 0,
%	\label{eqn:energy_eqn_full}
%\end{equation}
%where $\bm{F}_{\text{cond}} = -\kappa \grad T$.  
Understanding these flux terms and how they interact with the conductive flux, 
$\bm{F}_{\text{cond}} = -\kappa \grad T$, is crucial in characterizing
convective heat transport.

The atmosphere is contained between two impenetrable, stress free, fixed temperature boundaries at
the top and bottom of the domain such that $w = \partial_z u = T_1 = 0$ at $z = \{0, L_z\}$. The domain
is horizontally periodic. We utilize the novel Dedalus\footnote{http://dedalus-project.org/} pseudospectral framework 
 to time-evolve Eqs. 
\ref{eqn:continuity_eqn}-\ref{eqn:energy_eqn} using an implicit-explicit, third-order, four-step 
Runge-Kutta timestepping scheme RK443 \cite{ascher&all1997}.  
Variables are time-evolved on a dealiased Chebyshev (vertical)
and Fourier (horizontal) domain in which the
physical grid dimensions are 3/2 the size of the coefficient grid.  Physical grid sizes range from
96x384 grid points at the lowest values of Ra to 1152x4608 grid points at Ra $\geq 10^{7}$. 
By using IMEX timestepping, we implicitly step the stiff linear acoustic wave contribution and we are able to
efficiently study flows at moderate ($\text{Ma } \approx 1$) and very low ($\text{Ma } \approx 10^{-4}$)
Mach number (Fig. \ref{fig:ma_v_eps}b).  Our equations take the form
of the FC equations in \cite{lecoanet&all2014}, extended to include variable
$\nu$ and $\chi$, and we follow the approach there; this IMEX approach has been successfully 
tested against a nonlinear benchmark  of the compressible Kelvin-Helmholtz instability \cite{Lecoanet_et_al_2016_KH}.

>>>>

\section{Experiment} 
\refstepcounter{section}
\label{sec:experiment}
In order to compare our results with previous studies and in an effort to examine a simplest case,
we study a fluid composed of monatomic ideal gas particles with an adiabatic index of $\gamma = 5/3$ and
whose equation of state is $P = R\rho T$. 
The initial atmosphere is a plane-parallel polytrope in which the gravitational acceleration
and conductive flux, $\bm{F}_{\text{cond,0}} = -\kappa \grad T_0$, do not vary with depth. To
achieve the latter condition, both $\kappa$ and $\grad T_0$ are constant.
Under these assumptions, satisfying hydrostatic equilibrium produces a stratification of
\begin{equation}
\begin{split}
\rho_0(z) &= \rho_{t}(z_0 - z)^m, \\
T_0(z)    &= T_{t}(z_0 - z),
\label{eqn:polytrope}
\end{split}
\end{equation}
where $m = m_{ad} - \epsilon$ is the polytropic index.
The adiabatic polytropic index is $m_{ad} \equiv (\gamma-1)^{-1}$ and
the superadiabatic excess is $\epsilon$ which sets the scale 
of the entropy gradient, $\grad S_0 \propto -\epsilon$.
Thermodynamic variables are nondimensionalized at the top of the atmosphere as 
$P_0(L_z) = \rho_0(L_z) = T_0(L_z) = 1$, requiring $z_0 \equiv L_z + 1$ and $R = T_{t} = \rho_{t} = 1$.
By this choice, the non-dimensional length scale is the inverse temperature gradient scale and
the timescale is the isothermal sound crossing time of this unit length.
$z$ increases upwards within $[0, L_{z}]$, where $L_{z} = e^{n_{\rho}/m} - 1$ is
determined by $n_\rho$, the number of density scale heights the atmosphere spans.
The characteristic timescale of convective dynamics
is related to the atmospheric buoyancy time, $t_{\text{b}} = \sqrt{L_z/g\epsilon}$, with $g = (m+1)$.
Throughout this letter, buoyancy time units are used and 
$n_{\rho} = 3$ such that the initial density varies by a factor of 20.
All atmospheres studied here have an aspect ratio of 4, such that $L_x = 4L_z$.

Atmospheric diffusivities are specified by the Rayleigh number and the Prandtl number.  The
non-dimensional Rayleigh number is
\begin{equation}
\text{Ra} = \frac{g L_z^3 (\Delta S_0 / c_P)}{\nu\chi},
\end{equation}
where $\Delta S_0$ is the entropy difference between the top and bottom of the atmosphere, 
$c_P = R\gamma(\gamma-1)^{-1}$ is the specific heat at constant pressure,
$\nu$ is the kinematic viscosity (viscous diffusivity), and $\chi$ is the thermal diffusivity.  
The relationship between the thermal and viscous diffusivities is
set by the Prandtl number, $\text{Pr} = \nu/\chi$.   We relate the dynamic viscosity, $\mu$, and the thermal conductivity,
$\kappa$, to their corresponding diffusivities such that 
$\nu \equiv \mu/\rho$ and $\chi \equiv \kappa/\rho$.  As a result, $\text{Ra} \propto (\nu\chi)^{-1} \propto
\rho^2$.  The atmospheres studied here with $n_{\rho} = 3$ experience an increase in the Rayleigh number 
by a factor of 400 across the domain.  This formulation leaves Pr
constant throughout the depth of the atmosphere. In this letter we impose $\text{Pr} = 1$ and
specify Ra at $z = L_z$.

@@

The Fully Compressible Navier-Stokes equations,
\begin{align}
&\begin{aligned}
&\frac{\partial \ln\rho}{\partial t} + \Div{\bm{u}} = -\bm{u}\cdot\grad\rho,
	\label{eqn:continuity_eqn}
\end{aligned}\\
&\begin{aligned}
\frac{\partial\bm{u}}{\partial t} + \grad T - &\nu\Div{\lilstressT} - \lilstressT\cdot\grad\nu = \\
&-\bm{u}\cdot\grad\bm{u} - T\grad\ln\rho + \bm{g} + \nu\lilstressT\cdot\grad\ln\rho,
\label{eqn:momentum_eqn}
\end{aligned}\\
&\begin{aligned}
\frac{\partial T}{\partial t} -\frac{1}{c_V}\left(\right.\chi&\left.\grad^2 T + \grad T\cdot\grad\chi\right) = \\
	&-\bm{u}\cdot\grad T - (\gamma-1)T\Div{\bm{u}} \\
	&+ \frac{1}{c_V}\left(\chi\grad T \cdot\grad\ln\rho +\nu\left[\lilstressT\cdot\nabla\right]\cdot\bm{u}\right), 
	\label{eqn:energy_eqn}
\end{aligned}
\end{align}
are evolved with a viscous stress tensor defined as
\begin{equation}
\sigma_{ij} \equiv \left(\frac{\partial u_i}{\partial x_j} + \frac{\partial u_j}{\partial x_i} - \frac{2}{3}\delta_{ij}\Div{\bm{u}}\right).
	\label{eqn:stress_tensor}
\end{equation}
These equations carry a total convective flux of the form
%\begin{equation}
$
\bm{F}_{\text{conv}} \equiv \bm{F}_{\text{enth}} + \bm{F}_{\text{KE}} + \bm{F}_{\text{PE}} + \bm{F}_{\text{visc}}.
$
%\end{equation}
Contributions from the enthalpy flux, $\bm{F}_{\text{enth}} \equiv \rho\bm{u}(c_V T + P/\rho)$;
the kinetic energy flux, $\bm{F}_{\text{KE}} \equiv \rho|\bm{u}|^2\bm{u}/2$;
the potential energy flux, $\bm{F}_{\text{PE}} \equiv \rho\bm{u}\phi$ (with $\phi \equiv -gz$);
and the viscous flux, $\bm{F}_{\text{visc}} \equiv -\rho\nu\bm{u}\cdot\lilstressT$, must all be considered. 
%Taking an inner product of
%Eq. \ref{eqn:momentum_eqn} with $\bm{u}$ and adding it to 
%Eq. \ref{eqn:energy_eqn}, the full energy equation in conservation form is retrieved,
%\begin{equation}
%\frac{\partial}{\partial t}\left(\rho\left[\frac{|\bm{u}|^2}{2} + c_V T + \phi\right]\right) +
%\Div{\bm{F}_{\text{conv}} + \bm{F}_{\text{cond}}} = 0,
%	\label{eqn:energy_eqn_full}
%\end{equation}
%where $\bm{F}_{\text{cond}} = -\kappa \grad T$.  
Understanding these flux terms and how they interact with the conductive flux, 
$\bm{F}_{\text{cond}} = -\kappa \grad T$, is crucial in characterizing
convective heat transport.

The atmosphere is contained between two impenetrable, stress free, fixed temperature boundaries at
the top and bottom of the domain such that $w = \partial_z u = T_1 = 0$ at $z = \{0, L_z\}$. The domain
is horizontally periodic. We utilize the novel Dedalus\footnote{http://dedalus-project.org/} pseudospectral framework 
 to time-evolve Eqs. 
\ref{eqn:continuity_eqn}-\ref{eqn:energy_eqn} using an implicit-explicit, third-order, four-step 
Runge-Kutta timestepping scheme RK443 \cite{ascher&all1997}.  
Variables are time-evolved on a dealiased Chebyshev (vertical)
and Fourier (horizontal) domain in which the
physical grid dimensions are 3/2 the size of the coefficient grid.  Physical grid sizes range from
96x384 grid points at the lowest values of Ra to 1152x4608 grid points at Ra $\geq 10^{7}$. 
By using IMEX timestepping, we implicitly step the stiff linear acoustic wave contribution and are able to
efficiently study flows at moderate ($\approx 1$) and very low ($\approx 10^{-4}$)
Ma (Fig. \ref{fig:ma_v_eps}b).  Our equations take the form
of the FC equations in \cite{lecoanet&all2014}, extended to include variable
$\nu$ and $\chi$, and we follow the approach there; this IMEX approach has been successfully 
tested against a nonlinear benchmark  of the compressible Kelvin-Helmholtz instability \cite{Lecoanet_et_al_2016_KH}.
